---
title: "Exercise Analysis"
author: "Jillian Katz"
date: "2/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intoduction

## Project for Coursera (Hopkins) Practical Machne Learning

### Jillian Katz

The purpose of this project is to analyze accelerometer data to determine if fitness accelerometer wearers are performing set exercises correctly or incorrectly.  The study data set includes 6 participants performing exercises in 5 different incorrect & correct ways (classified as A, B, C, D, & E.

Thank you to [Groupware](http://groupware.les.inf.puc-rio.br/har) for supplying this data! 

### Note on Assignment:
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 


```{r}
library(readr)
pml_training <- read_csv("C:/Users/jkjil/Downloads/pml-training.csv")
View(pml_training)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
