---
title: "Regression Project"
author: "Jillian Katz"
date: "1/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## J Katz

## Coursera Regression Course Project

### Introduction

Here I used the Motor Trend Car Road Tests dataset to analyze if - or how much - difference in car miles per gallon (MPG) can be attributed to the type of transmission (automatic vs manual).

### Analysis

To begin the analysis, I compared automatic and manual transmission cars' MPG independent of other variables, and there is an apparent difference.  The average gas mileage for a Manual Transmission vehicle is about 24 mpg, compared to about 17 mpg for an Automatic.  A Student's T test verified that the data shows a difference in the transmission types with over 99% confidence.

```{r, echo = F, results = 'hide'}
library(ggplot2)
library(dplyr)
```

```{r, echo = F}
ggplot(data = mtcars, aes(x =am,y = mpg, group = am, 
                          fill = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_boxplot() + labs(fill = "Transmission") +
  xlab("Transmission") + ylab("MPG")

autompg <- filter(mtcars, am == 0)$mpg
manmpg <- filter(mtcars, am == 1)$mpg
t <-t.test(autompg, manmpg, alternative = "two.sided")
paste("T Test P value:", round(t$p.value, 6))

# t = -3.7671, df = 18.332, p-value = 0.001374
```


However, there may be a confounding influence from other variables in the data.  A linear model predicting MPG with all of the other variables[^1]  showed an expected increase in 2.8mpg if changing from Automatic to Manual transmission.  This value came with a p-value of 0.16, which is not necessarily significant.

[^1]: Variables in the dataset: Miles per gallon, Number of cylinders, Displacement, Horsepower, Rear axle ratio, Weight, Quarter mile time, Engine shape, Transmission, Number of forward gears, & Number of carburetors

I created additional models to find which variables truly have the strongest influence on MPG.  I landed on transmission, rear axle ratio, weight, and quarter mile time.  A linear model with these factors showed weight, quarter mile time, and transmission to be significant (p-values < 0.1).  Overall, this suggests that the difference between a Manual and Automatic transmission vehicle results in a gain of about 3 miles per gallon.  

Notably, there is an opposite influence by weight: A  vehicle weight gain of 1000 lb suggests a loss of about 3mpg.  Automatic vehicles tend to be heavier as well. This correlation explains the difference between effect of transmission type when the model excludes weight. 

While this model is not perfect, it does fit the data reasonably well.  The plot of the residuals shows a slight curve, but no major patterns.


```{r, echo = F}
fit5 <- lm(mpg ~ drat+ wt + qsec + am -1, data = mtcars)
#summary(fit5)
plot(fit5, 1)

```


### Summary
The 1974 Motor Trend data suggests that manual transmission vehicles do get superior gas mileage compared to automatic transmission vehicles.  However, the difference is only about 3mpg - considering other factors in cars - compared to the difference of 7mpg - when not considering other variables.

## Additional Plots

```{r, echo=F, results = 'hide'}

# wt vs mpg - auto cars are heavier
ggplot(data = mtcars, aes(x = wt , y = mpg, color = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_point() + labs(color = "Transmission") +
  xlab("Weight") + ylab("MPG") + ggtitle("Automatic cars tend to be heavier")

# cyl vs mpg - auto have more cylinders
ggplot(data = mtcars, aes(x = cyl , y = mpg, color = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_point() + labs(color = "Transmission") +
  xlab("Cylinders") + ylab("MPG") + ggtitle("Fewer cylinders: better gas mileage and manual transmission.")
# hp vs mpg - slight effect?
ggplot(data = mtcars, aes(x = hp , y = mpg, color = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_point() + labs(color = "Transmission") +
  xlab("Horsepower") + ylab("MPG") + ggtitle("More manual cars have lower horsepower.")
#qsec - clear difference in performance
ggplot(data = mtcars, aes(x =qsec , y = mpg, color = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_point() + labs(color = "Transmission")+
  xlab("Quarter Mile Time") + ylab("MPG") + ggtitle("Automatic cars might accelerate slightly faster.")

#rear axle ratio - clear difference in performance
ggplot(data = mtcars, aes(x =drat , y = mpg, color = factor(am, labels = 
                                          c("Automatic", "Manual")))) +
  geom_point() + labs(color = "Transmission")+
  xlab("Rear Axle Ratio") + ylab("MPG") + ggtitle("Manual cars tend to have higher rear axle ratios.")

plot(fit5)
```

## The Code

```{r, echo=T, results = 'hide'}
library(ggplot2)
library(dplyr)

## t test mpg vs transmission
autompg <- filter(mtcars, am == 0)$mpg
mean(autompg) # 17.14737
manmpg <- filter(mtcars, am == 1)$mpg
mean(manmpg) # 24.39231
t.test(autompg, manmpg, alternative = "two.sided")
# t = -3.7671, df = 18.332, p-value = 0.001374


fit <- lm(mpg ~. -1, data = mtcars)
summary(fit) ##  p val = .2340, low effect of transmission when including all variables. smallest ps from wt and qsec
plot(fit)# residuals vs fitted look OK- no patterns


fit2 <- lm(mpg ~ wt + qsec + am -1, data = mtcars)
summary(fit2) # all factors significant
plot(fit2)

fit3 <- lm(mpg ~ wt + am -1, data = mtcars) # does transmission matter if considering weight?
summary(fit3) # yes
plot(fit3)

# fit for all with coeff > 1
fit4 <- lm(mpg ~ drat+ wt + qsec + am + gear -1, data = mtcars)
summary(fit4)#gear has very low coeff, p almost 1
plot(fit4)

# fit4 with gear removed
fit5 <- lm(mpg ~ drat+ wt + qsec + am -1, data = mtcars)
summary(fit5)
plot(fit5)

fit6 <- lm(mpg ~ drat + qsec + am -1, data = mtcars)
summary(fit6) # effect of transmission without weight: about the same as the mean difference of automatic and manual
plot(fit6)

## Exploration of relationships
# transmission vs mpg - manual (1) looks better
ggplot(data = mtcars, aes(x = am , y = mpg, color = am)) + geom_point()

# wt vs mpg - auto cars are heavier
ggplot(data = mtcars, aes(x = wt , y = mpg, color = am)) + geom_point()

# cyl vs mpg - auto have more cylinders
ggplot(data = mtcars, aes(x = cyl , y = mpg, color = am)) + geom_point()
# hp vs mpg - slight effect
ggplot(data = mtcars, aes(x = hp , y = mpg, color = am)) + geom_point()
#qsec
ggplot(data = mtcars, aes(x =qsec , y = mpg, color = am)) + geom_point()


```
